---
title: "30 day rolling average analysis"
format: pdf
---

```{r}
library(ggplot2)
library(mgcv)
library(dplyr)
library(RColorBrewer)

df.ts.series2 <- read.csv("df.ts.series2.csv") %>% 
  mutate(tree_index = as.character(tree_index)) %>%
  filter(!species == "Pseudotsuga menziesii")
num.cols <- c("twd", "pr", "at", "ws", "dp", "sr", "lr", "day.of.year")

year_marks <- c(0, 366, 731, 1096) # 2020 is a leap year
xlab <- "Cumulative Time (days)"
``` 


```{r}
rolling_avg_cols <- colnames(df.ts.series2)[grep("_rolling30", colnames(df.ts.series2))]

rolling_avg_cols <- rolling_avg_cols[!rolling_avg_cols =="day.of.year_rolling30"]

print_time_ycol <- function(ycol){
  ggplot(
    df.ts.series2,
    aes(x = cum.time.days, color = year, group = interaction(year, tree.id))) +
    geom_line(aes_string(y = ycol)) +
    #geom_point(aes(color = year), size = 1)+
    facet_wrap(~species) +
    theme_bw() +
    theme(legend.position='bottom')
}
```

```{r fig.height = 6}

df.ts.series2.plot <- df.ts.series2
nb_indexes <- length(unique(df.ts.series2.plot$tree_index))

vec1 <- brewer.pal(n = 12, name = "Paired")
vec2 <- brewer.pal(n = nb_indexes-12, name = "Paired")
vec_color <- c(vec1, vec2)

bands <- read.csv("bands_seasonal.csv")

get_plot_basic <-  function(my_df, my_y_col, my_y_title ){
  plot_basic <- ggplot(my_df, aes(x = cum.time.days, color = tree_index, group = tree.id)) +
    geom_line(aes_string(y = my_y_col)) +
    theme_bw() +
    theme(legend.position='none') +
    guides(color = guide_legend(nrow = 3)) +
    labs(
      x = xlab, y= my_y_title,
      color = "Tree index within each species") +
    scale_color_manual(values = vec_color)
}

add_details  <-  function(my_plot){
  my_plot +
    geom_vline(xintercept = year_marks, linetype = "dotted") +
    geom_rect(
      data = bands,
      aes(xmin = xmin, xmax = xmax, ymin = -Inf, ymax = Inf, fill = season),
      inherit.aes = FALSE, alpha = 0.12
    ) +
    scale_fill_manual(values = c(Winter = "grey50", Summer = "goldenrod")) +
    facet_wrap(~species, ncol = 2) 
}

plots_rolling <- lapply(rolling_avg_cols, get_plot_basic, my_df = df.ts.series2.plot, my_y_title = "30-day Rolling average TWD (mm)")
plots_rolling2 <- lapply(plots_rolling, add_details)
plot_time_twd_rolling_avg <- plots_rolling2[[1]]
ggsave(
  filename= "images for report/plot_time_twd_rolling_avg.png", 
  plot= plot_time_twd_rolling_avg, height = 10, width = 8)
``` 


```{r}
rolling_avg_cols_7 <- colnames(df.ts.series2)[grep("_rolling7", colnames(df.ts.series2))]
plots_rolling_7 <- lapply(rolling_avg_cols_7, get_plot_basic, my_df = df.ts.series2.plot, my_y_title = "7-day Rolling average TWD (mm)")
plots_rolling_7 <- lapply(plots_rolling_7, add_details)
plot_time_twd_rolling_avg_7 <- plots_rolling_7[[1]]
ggsave(
  filename= "images for report/plot_time_twd_rolling_avg_7.png", 
  plot= plot_time_twd_rolling_avg_7, height = 10, width = 8)
``` 

```{r rolling_SD}
rolling_avg_cols_SD <- colnames(df.ts.series2)[grep("_SDrolling30", colnames(df.ts.series2))]
plots_rolling_SD <- lapply(rolling_avg_cols_SD, get_plot_basic, my_df = df.ts.series2.plot, my_y_title = "30-day Rolling SD TWD (mm)")
plots_rolling_SD <- lapply(plots_rolling_SD, add_details)
plot_time_twd_rolling_SD <- plots_rolling_SD[[1]]
ggsave(
  filename= "images for report/plot_time_twd_rolling_SD.png", 
  plot= plot_time_twd_rolling_SD, height = 10, width = 8)
```

```{r}
rolling_avg_cols
# s(cum.time.days)
# Create formula for GAM in a simpler way
predictors <- rolling_avg_cols[rolling_avg_cols != "twd_rolling30"]
smooth_terms <- paste0("s(", predictors, ")", collapse = " + ")
formula_str <- paste("twd_rolling30 ~", smooth_terms, "+ species")
gam_model <- gam(as.formula(formula_str), data = df.ts.series2)
```

#  Partial effect plots of fitted Generalized Additive Model (GAM)

```{r fig.height = 5}
anova(gam_model)
summary(gam_model)
plot(gam_model, pages = 1, se = TRUE)
``` 

General structure of each plot:

- X-axis: The values of the predictor variable (e.g., pr_rolling30, at_rolling30, etc.).

- Y-axis: The smooth effect f(x) of the predictor on the response. This is not the raw coefficient, but the estimated non-linear transformation.

- Black line: Estimated smooth function.

- Dashed lines: Confidence intervals (usually 95%).

Individual variable interpretations:

pr_rolling30 (precipitation?): Effect is fairly flat → very little to no influence on the response in this range.

at_rolling30 (air temperature?): U-shaped effect → lower values are associated with higher outcomes, middle values suppress it, and higher values raise it again.

ws_rolling30 (wind speed?): Slightly declining influence, but very flat → weak or no effect overall.

dp_rolling30 (dew point?): Positive effect when very low, then peaks and slightly drops → might indicate an optimal dew point range.

sr_rolling30 (solar radiation?): Flat and near-zero → suggests solar radiation has little effect on the response.

lr_rolling30 (longwave radiation?): Slight decline → higher values may slightly decrease the response.


