---
title: "30 day rolling average analysis"
format: pdf
---

```{r}
library(ggplot2)
library(mgcv)
library(dplyr)

df.ts.series2 <- read.csv("df.ts.series2.csv") %>% 
  filter(!species == "Pseudotsuga menziesii")
num.cols <- c("twd", "pr", "at", "ws", "dp", "sr", "lr", "day.of.year")
``` 

```{r}
rolling_avg_cols <- colnames(df.ts.series2)[grep("_rolling30", colnames(df.ts.series2))]

rolling_avg_cols <- rolling_avg_cols[!rolling_avg_cols =="day.of.year_rolling30"]

print_time_ycol <- function(ycol){
  ggplot(
    df.ts.series2,
    aes(x = cum.time.days, color = year, group = interaction(year, tree.id))) +
    geom_line(aes_string(y = ycol)) +
    #geom_point(aes(color = year), size = 1)+
    facet_wrap(~species) +
    theme(legend.position='bottom')
}
``` 

```{r fig.height = 6}
lapply(rolling_avg_cols, print_time_ycol)
``` 

```{r}
rolling_avg_cols
df.ts.series2$year <- factor(df.ts.series2$year)
gam_model <- gam(twd_rolling30 ~ s(cum.time.days) +  s(pr_rolling30) + s(at_rolling30) + s(ws_rolling30) + s(dp_rolling30) + s(sr_rolling30) + s(lr_rolling30) + year + species, data = df.ts.series2)
```

#  Partial effect plots of fitted Generalized Additive Model (GAM)

```{r fig.height = 10}
anova(gam_model)
summary(gam_model)
plot(gam_model, pages = 1, se = TRUE)
``` 

General structure of each plot:

- X-axis: The values of the predictor variable (e.g., pr_rolling30, at_rolling30, etc.).

- Y-axis: The smooth effect f(x) of the predictor on the response. This is not the raw coefficient, but the estimated non-linear transformation.

- Black line: Estimated smooth function.

- Dashed lines: Confidence intervals (usually 95%).

Individual variable interpretations:

pr_rolling30 (precipitation?): Effect is fairly flat → very little to no influence on the response in this range.

at_rolling30 (air temperature?): U-shaped effect → lower values are associated with higher outcomes, middle values suppress it, and higher values raise it again.

ws_rolling30 (wind speed?): Slightly declining influence, but very flat → weak or no effect overall.

dp_rolling30 (dew point?): Positive effect when very low, then peaks and slightly drops → might indicate an optimal dew point range.

sr_rolling30 (solar radiation?): Flat and near-zero → suggests solar radiation has little effect on the response.

lr_rolling30 (longwave radiation?): Slight decline → higher values may slightly decrease the response.


