---
title: "Tree water deficit: data exploration"
author: Salomé Santos
format: pdf
---

```{r load_libraries, message = FALSE, warning = FALSE}
require(dplyr)
require(ggplot2)
require(car)
require(stringr)
require(lubridate)
require(zoo)
#require(flextable)
library(broom)
library(ggplot2)
library(forecast)

knitr::opts_chunk$set(fig.width=7, fig.height=7) 
```

```{r}
bool_run_dataset <- TRUE

if(bool_run_dataset){source("read.datasets.r")
  } else {
  df.ts.series2 <- read.csv("df.ts.series2.csv")
  num.cols <- c("twd", "pr", "at", "ws", "dp", "sr", "lr", "day.of.year")
  }
``` 


```{r fig.height = 10}

plot_basic <- ggplot(df.ts.series2, aes(x = cum.time.days, y = twd, group = tree.id)) +
  geom_line(aes(color = tree.id)) +
  theme_bw() +
  theme(legend.position='bottom') +
  guides(color = guide_legend(nrow = 3))

plot_basic +
  facet_grid(vars(species), vars(site))

plot_basic +
  geom_vline(xintercept = 365, color = "blue") +
  geom_vline(xintercept = 730, color = "blue") +
  facet_wrap(~species)

ggplot(
  df.ts.series2,
  aes(x = day.of.year, y = twd,
  group = interaction(year, tree.id))) +
  geom_line(aes(color = year)) +
  facet_wrap(~species) +
  theme(legend.position='bottom')
```

```{r}

#lapply(df.ts.series2, class)

num.vars.df <- df.ts.series2[,num.cols]
cor_matrix <- cor(num.vars.df, use = "complete.obs")
#flextable(cor_matrix)
cor_matrix
#pairs(num.vars.df)
```

```{r corr_by_species}
list_species <- unique(df.ts.series2$species)

for(iter_species in list_species){
  iter_df <- df.ts.series2 %>%
    filter(species == iter_species)

  print(iter_species)
  print(cor(iter_df[,num.cols]))
}
```

```{r}
ggplot(
  df.ts.series2,
  aes(x = at, y = twd,
  group = interaction(year, tree.id))) +
  geom_point(aes(color = year)) +
  facet_wrap(~species) +
  theme(legend.position='bottom')
``` 

```{r fig.height = 10}
ggplot(
  df.ts.series2,
  aes(x = ts, y = twd_rolling7,
  group = interaction(year, tree.id))) +
  geom_line(aes(color = year)) +
  facet_wrap(~species) +
  theme(legend.position='bottom')

ggplot(
  df.ts.series2,
  aes(x = ts, y = twd_rolling30,
  group = interaction(year, tree.id))) +
  geom_line(aes(color = year)) +
  #geom_point(aes(color = year), size = 1)+
  facet_wrap(~species) +
  theme(legend.position='bottom')

ggplot(
  df.ts.series2,
  aes(x = at_rolling30, y = twd_rolling30,
  group = interaction(year, tree.id))) +
  geom_point(aes(color = year)) +
  facet_wrap(~species) +
  theme(legend.position='bottom')

ggplot(
  df.ts.series2,
  aes(x = pr_rolling30, y = twd_rolling30,
  group = interaction(year, tree.id))) +
  geom_point(aes(color = year)) +
  facet_wrap(~species) +
  theme(legend.position='bottom')  

acf(df.ts.series2$twd, lag.max = 500)
# clear cycl because high correlation again at 365 days

acf(df.ts.series2$at, lag.max = 500)
```

```{r corr_by_species_rolling7}
rolling.cols <- paste0(num.cols, "_rolling7")

for(iter_species in list_species){
  iter_df <- df.ts.series2 %>%
    filter(species == iter_species)

  print(iter_species)
  print(cor(iter_df[,rolling.cols], use = "complete.obs"))
}
```

```{r corr_by_species_rolling30}
rolling.cols <- paste0(num.cols, "_rolling30")

for(iter_species in list_species){
  iter_df <- df.ts.series2 %>%
    filter(species == iter_species)

  print(iter_species)
  print(cor(iter_df[,rolling.cols], use = "complete.obs"))
}
```

## Partial Autocorrelation
```{r}
# Plot PACF
ggPacf(df.ts.series2$twd)
ggPacf(df.ts.series2$twd_rolling30)
```


```{r}
plot_twd_lag <- function(df, x_col, twd_col, lag = 1) {

  # Arrange data by time
  df <- df %>% arrange(.data[[x_col]])

  # Create lagged TWD values
  df <- df %>%
    mutate(Lagged_TWD = lag(.data[[twd_col]], lag))

  # Plot TWD vs. lagged TWD
  ggplot(df, aes(x = ts, color = tree.id, y = .data[[twd_col]])) +
    geom_line() +
    geom_line(aes(y= Lagged_TWD))+
    labs(title = paste("Tree Water Deficit vs. Lagged TWD (Lag =", lag, ")"),
         x = paste("TWD at Lag", lag),
         y = "TWD at Current Time") +
    theme_bw() +
    facet_wrap(~species)
}

plot_twd_lag(df.ts.series2, "at", "twd", lag = 5)

```

# Granger Causality

Granger Causality: Rather than testing whether X causes Y, the Granger causality tests whether X forecasts Y. 

A time series X is said to Granger-cause Y if it can be shown, usually through a series of t-tests and F-tests on lagged values of X (and with lagged values of Y also included), that those X values provide statistically significant information about future values of Y. 

A variable X that evolves over time Granger-causes another evolving variable Y if predictions of the value of Y based on its own past values and on the past values of X are better than predictions of Y based only on Y's own past values.  

If the variables are non-stationary, then the test is done using first (or higher) differences. The number of lags to be included is usually chosen using an information criterion


```{r message = FALSE, warning = FALSE}
library(vars)
library(urca)
library(vars)

summary(ur.df(df.ts.series2$twd, type="drift", lags=2))
summary(ur.df(df.ts.series2$at, type="drift", lags=2))
```

This tests the null hypothesis: The series has a unit root (i.e., non-stationary).

If the coefficient on z.lag.1 is significantly different from zero (with a very small p-value), then you can reject the null hypothesis and conclude the series is stationary.

```{r}

granger_test <- function(data, group = NA, x_col, y_col, lags = 10) {
  run_test <- function(df_sub) {
    df <- df_sub[, c(x_col, y_col)]
    names(df) <- c("x", "y")
    df <- df[complete.cases(df), ]
    
    if (nrow(df) <= lags + 1) {
      return("Not enough observations for VAR model.")
    }
    
    var_model <- VAR(df, p = lags, type = "const")
    return(causality(var_model, cause = "x"))
  }
  
  if (is.na(group)) {
    return(run_test(data))
  } else {
    group_list <- split(data, data[[group]])
    results <- lapply(group_list, run_test)
    return(results)
  }
}

granger_test(df.ts.series2, y_col= "twd", x_col= "at")
granger_test(df.ts.series2, y_col = "twd", x_col= "dp")

```

## Granger test by species

```{r}
granger_test(df.ts.series2, y_col= "twd", x_col= "at", group = "species")
```

## Linear regression

```{r}
plot_coeffs <- function(my_coef_df){
  ggplot(my_coef_df, aes(x = reorder(term, estimate), y = estimate)) +
    geom_bar(stat = "identity", fill = "steelblue") +
    geom_errorbar(aes(ymin = estimate - std.error, ymax = estimate + std.error), width = 0.2) +
    coord_flip() +  # flip axes for readability
    labs(
      title = "Linear Model Coefficients",
      x = "Predictor",
      y = "Estimate"
    ) +
    theme_minimal()
}
```

```{r}
lm_fit <- lm(
  formula = twd ~ cum.time.days +  at + dp + pr + ws + sr + lr,
  data = df.ts.series2
)

summary(lm_fit)
coef_df <- tidy(lm_fit)  # gets estimate, std.error, etc.
plot_coeffs(coef_df)
```  

```{r}
lm_fit <- lm(
  formula = twd ~ cum.time.days + at + dp + pr + ws + sr + lr,
  data = df.ts.series2,
  subset = species %in% c("Picea abies")
)

summary(lm_fit)
coef_df <- tidy(lm_fit)  # gets estimate, std.error, etc.
plot_coeffs(coef_df)
```  

```{r}
lm_fit <- lm(
  formula = twd ~ cum.time.days +  at + dp + pr + ws + sr + lr,
  data = df.ts.series2,
  subset = species %in% c("Fagus sylvatica")
)

summary(lm_fit)
coef_df <- tidy(lm_fit)  # gets estimate, std.error, etc.
plot_coeffs(coef_df)
```


# Rollins SD

Confirm that timeseries are not stationary

```{r fig.height =10}
library(dplyr)
library(purrr)
library(ggplot2)
library(broom)
library(tidyr)
library(ggplot2)

# Nest data by tree.id
nested <- df.ts.series2 %>%
  group_by(tree.id) %>%
  nest()

# Fit model and extract residuals
residuals_by_tree <- nested %>%
  mutate(
    model = map(data, ~ lm(twd ~ cum.time.days, data = .x)),
    augmented = map(model, augment)
  ) %>%
  unnest(augmented)

residuals_by_tree_p1 <- residuals_by_tree %>% 
  filter(tree.id < 22)
residuals_by_tree_p2 <- residuals_by_tree %>% 
  filter(tree.id > 21)

plot_residuals <- function(df){
  # Plot residuals vs fitted for each tree
  print(ggplot(df, aes(.fitted, .resid)) +
    geom_point(alpha = 0.6) +
    geom_hline(yintercept = 0, color = "red") +
    facet_wrap(~ tree.id, scales = "free", ncol = 4) +
    labs(x = "Fitted values", y = "Residuals",
        title = "Residuals vs Fitted Values for Each Tree ID") +
    theme_minimal())

  print(ggplot(df, aes(x = cum.time.days, y = .resid)) +
    geom_point(alpha = 0.6) +
    geom_hline(yintercept = 0, color = "red") +
    facet_wrap(~ tree.id, scales = "free", ncol = 4) +
    labs(x = "Cumulative Time (days)", y = "Residuals",
        title = "Residuals vs. Cumulative Time by Tree ID") +
    theme_minimal())
}

plot_residuals(residuals_by_tree_p1)
plot_residuals(residuals_by_tree_p2)

# Step 1: Fit a linear model for each tree
coeffs_by_tree <- df.ts.series2 %>%
  group_by(tree.id) %>%
  nest() %>%
  mutate(
    model = map(data, ~ lm(twd ~ cum.time.days, data = .x)),
    coef = map(model, tidy)
  ) %>%
  unnest(coef)

# Step 2: Plot coefficients
ggplot(coeffs_by_tree, aes(x = tree.id, y = estimate, fill = term)) +
  geom_bar(stat = "identity", position = position_dodge()) +
  labs(x = "Tree ID", y = "Coefficient Estimate", fill = "Term",
       title = "Linear Model Coefficients by Tree ID") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 90))
```