---
title: "Tree water deficit: data exploration"
author: Salomé Santos
format: pdf
---

```{r load_libraries, message = FALSE, warning = FALSE}
require(dplyr)
require(ggplot2)
require(car)
require(stringr)
require(lubridate)
require(zoo)
#require(flextable)
library(broom)
library(ggplot2)
library(forecast)
library(RColorBrewer)
library(purrr)
library(tidyr)
library(broom)
library(gridExtra)

knitr::opts_chunk$set(fig.width=7, fig.height=7) 
```

```{r}
bool_run_dataset <- TRUE

if(bool_run_dataset){source("read.datasets.r")
  } else {
  df.ts.series2 <- read.csv("df.ts.series2.csv")
  num.cols <- c("twd", "pr", "at", "ws", "dp", "sr", "lr", "day.of.year")
  }
df.ts.series2$tree_index <- as.factor(df.ts.series2$tree_index)
``` 


```{r fig.height = 10}
# for the report

 # Pick one random tree from df_plot_fagus
set.seed(123) # for reproducibility
species1_tree_ids <- unique(df.ts.series$tree.id[df.ts.series$species == "Carpinus betulus"])
random_tree_id_species1 <- sample(species1_tree_ids, 1)

species2_tree_ids <- unique(df.ts.series$tree.id[df.ts.series$species == "Picea abies"])
random_tree_id_species2 <- sample(species2_tree_ids, 1)

df_plot_certain_species <- df.ts.series %>%
  filter(tree.id %in% c(random_tree_id_species1, random_tree_id_species2 )) %>%
  group_by(tree.id) %>%
  mutate(
    across(
      c(twd, at, pr, lr, sr, ws, dp),
      ~ (. - mean(., na.rm = TRUE)) / sd(., na.rm = TRUE)
    )
  )

year_marks <- c(0, 366, 731, 1096) # 2020 is a leap year
xlab <- "Cumulative Time (days)"

get_plot_twd_and_covariate <- function(df, my_covariate) {
  
  p <- ggplot(df, aes(x = cum.time.days)) +
    geom_line(aes(y = twd), color = "#00008B") +
    geom_line(aes_string(y = my_covariate), alpha = 0.5, color = "red") +
    geom_vline(xintercept = year_marks, linetype = "dotted") +
    facet_wrap(~ species, ncol = 2) +
    theme_bw() +
    labs(
      x = xlab,
      y = paste("Standardized Values of TWD (dark blue) and", toupper(my_covariate), "(red)"),
      color = "Variable"
    ) +
    theme(legend.position = "bottom")
  
  return(p)
}

time_vs_twd_at <- get_plot_twd_and_covariate(df_plot_certain_species, my_covariate = "at")
ggsave("images for report/time_vs_twd_at.png", plot = time_vs_twd_at, width = 10, height = 5)
time_vs_twd_pr <- get_plot_twd_and_covariate(df_plot_certain_species, my_covariate = "pr")
ggsave("images for report/time_vs_twd_pr.png", plot = time_vs_twd_pr, width = 10, height = 5)

vec_covariates <- c("at", "pr", "lr", "sr", "ws", "dp")

for(iter_cov in vec_covariates) {
  get_plot_twd_and_covariate(df_plot_certain_species, my_covariate = iter_cov)
  ggsave(filename = paste0("images for report/time_vs_twd_", iter_cov, ".png"), width = 10, height = 5)
}
```

```{r create_season_bands}

# Summer = Jun 1 (DOY 152) to Aug 31 (DOY 243)
summer_doy <- c(152, 243)

# Winter = Dec 1 (DOY 335) to Dec 31 (365), AND Jan 1 (1) to Feb 28/29 (59)
winter1_doy <- c(335, 365)  # Dec
winter2_doy <- c(1,   59)   # Jan–Feb

# If cum.time.days starts at 0
origin_shift <- 0
min(df.ts.series2$cum.time.days)

# X range in your data
xr <- range(df.ts.series2.plot$cum.time.days, na.rm = TRUE)

# Ensure year_marks cover the whole range (add end mark if needed)
ym <- sort(unique(year_marks))
if (max(ym) < xr[2]) ym <- c(ym, ceiling(xr[2] / 365) * 365)

# Helper to make one band given a year's start
make_band <- function(y0, doy) {
  tibble(
    xmin = y0 + (doy[1] - 1 + origin_shift),
    xmax = y0 + (doy[2] - 1 + origin_shift)
  )
}

# Build bands for each year segment
bands <- bind_rows(lapply(ym, function(y0) {
  # Summer (single span within the year)
  s <- make_band(y0, summer_doy) %>% mutate(season = "Summer")

  # Winter spans two chunks: Dec (within this year) + Jan–Feb (at next year's start)
  w_dec <- make_band(y0, winter1_doy) %>% mutate(season = "Winter")
  w_jf  <- make_band(y0 + 365, winter2_doy) %>% mutate(season = "Winter")

  bind_rows(s, w_dec, w_jf)
})) %>%
  # Keep only bands overlapping the plotting range
  filter(xmax >= xr[1], xmin <= xr[2])

write.csv(bands, "bands_seasonal.csv", row.names = FALSE)

```

```{r fig.height = 10}

nb_indexes <- length(unique( df.ts.series2$tree_index))

vec1 <- brewer.pal(n = 12, name = "Paired")
vec2 <- brewer.pal(n = nb_indexes-12, name = "Paired")
vec_color <- c(vec1, vec2)

df.Pseudotsuga <- df.ts.series2 %>%
  filter(species == "Pseudotsuga menziesii")
df.ts.series2 <- setdiff(df.ts.series2, df.Pseudotsuga)

get_plot_basic <-  function(my_df){
  plot_basic <- ggplot(my_df, aes(x = cum.time.days, y = twd, group = tree.id)) +
    geom_line(aes(color = tree_index)) +
    theme_bw() +
    theme(legend.position='none') +
    guides(color = guide_legend(nrow = 3)) +
    labs(
      x = xlab, y= "TWD (mm)",
      color = "Tree index within each species") +
    scale_color_manual(values = vec_color)
}
plot_basic <- get_plot_basic(df.ts.series2)

plot_basic +
  facet_grid(vars(species), vars(site))

add_details  <-  function(my_plot){
  my_plot +
    geom_vline(xintercept = year_marks, linetype = "dotted") +
    geom_rect(
      data = bands,
      aes(xmin = xmin, xmax = xmax, ymin = -Inf, ymax = Inf, fill = season),
      inherit.aes = FALSE, alpha = 0.12
    ) +
    scale_fill_manual(values = c(Winter = "grey50", Summer = "goldenrod")) +
    facet_wrap(~species, ncol = 2) 
}
plot_time_twd_by_species <- add_details(plot_basic )

ggsave(
  filename= "images for report/plot_time_twd_by_species.png", 
  plot= plot_time_twd_by_species, height = 10, width = 8) 


plot_Pseudotsuga <- add_details(get_plot_basic(df.Pseudotsuga))
ggsave(
  filename= "images for report/plot_time_twd_Pseudotsuga.png", 
  plot= plot_Pseudotsuga, height = 5, width = 7) 

ggplot(
  df.ts.series2,
  aes(x = day.of.year, y = twd,
  group = interaction(year, tree.id))) +
  geom_line(aes(color = year)) +
  facet_wrap(~species) +
  theme(legend.position='bottom')
```

```{r}

df_long <- df.ts.series2 %>%
  select(all_of(c(vec_covariates, "twd"))) %>%
  pivot_longer(cols = everything(), names_to = "variable", values_to = "value")

hist_unprocessed_vars <- ggplot(df_long, aes(x = value)) +
  geom_histogram(bins = 30, fill = "steelblue", color = "white", alpha = 0.7) +
  facet_wrap(~ variable, scales = "free", ncol =2) +
  theme_bw() +
  labs( x = "Value", y = "Count")
ggsave("images for report/hist_unprocessed_vars.png", plot = hist_unprocessed_vars, width = 8, height = 10)  
```

```{r}

#lapply(df.ts.series2, class)

num.vars.df <- df.ts.series2[,num.cols]
cor_matrix <- cor(num.vars.df, use = "complete.obs")
#flextable(cor_matrix)
cor_matrix
#pairs(num.vars.df)
```

```{r corr_by_species}
list_species <- unique(df.ts.series2$species)

for(iter_species in list_species){
  iter_df <- df.ts.series2 %>%
    filter(species == iter_species)

  print(iter_species)
  print(cor(iter_df[,num.cols]))
}
```

```{r}
ggplot(
  df.ts.series2,
  aes(x = at, y = twd,
  group = interaction(year, tree.id))) +
  geom_point(aes(color = year)) +
  facet_wrap(~species) +
  theme(legend.position='bottom')
``` 

```{r fig.height = 10}
ggplot(
  df.ts.series2,
  aes(x = ts, y = twd_rolling7,
  group = interaction(year, tree.id))) +
  geom_line(aes(color = year)) +
  facet_wrap(~species) +
  theme(legend.position='bottom')

ggplot(
  df.ts.series2,
  aes(x = ts, y = twd_rolling30,
  group = interaction(year, tree.id))) +
  geom_line(aes(color = year)) +
  #geom_point(aes(color = year), size = 1)+
  facet_wrap(~species) +
  theme(legend.position='bottom')

ggplot(
  df.ts.series2,
  aes(x = at_rolling30, y = twd_rolling30,
  group = interaction(year, tree.id))) +
  geom_point(aes(color = year)) +
  facet_wrap(~species) +
  theme(legend.position='bottom')

ggplot(
  df.ts.series2,
  aes(x = pr_rolling30, y = twd_rolling30,
  group = interaction(year, tree.id))) +
  geom_point(aes(color = year)) +
  facet_wrap(~species) +
  theme(legend.position='bottom')  

```

Missing values

```{r}
# Check which rows have missing TWD values
missing_twd_rows <- which(is.na(df.ts.series2$twd))
cat("Number of missing TWD values:", length(missing_twd_rows), "\n")
df.ts.series2 %>%
  filter(is.na(twd)) %>%
  select(tree.id, species) %>%
  distinct() 
df.ts.series2 <- df.ts.series2 %>% filter(!is.na(twd))
```

ACF

```{r}
unique_tree_ids <- unique(df.ts.series2$tree.id)

# Save ACF results for later plotting
get_acf <- function(tid, y_val = "twd") {
  ts_data <- df.ts.series2 %>% filter(tree.id == tid)
  acf_obj <- acf(ts_data[[y_val]], lag.max = 2000, plot = FALSE)

  df <- tibble(
    tree_index = unique(ts_data$tree_index),
    species= unique(ts_data$species),
    tree.id = tid,
    lag = acf_obj$lag[,1,1],
    acf = acf_obj$acf[,1,1]
  ) %>%
    filter(lag > 0)  # Exclude lag 0
   return(df)   
}

plot_corr_plots <- function(my_df, my_y_val = "acf", add_smooth = TRUE) {

  if(add_smooth){alpha = 0.5} else {alpha = 0.7}

  cacf_plot <- ggplot(
    my_df, 
    aes(x = lag, y = .data[[my_y_val]], color = tree_index)) +
      geom_point(alpha = alpha, size = 0.5, shape = 1) 

  if (add_smooth) {
    cacf_plot <- cacf_plot +
      geom_smooth(se = FALSE, size = 1, method = "loess", span = 0.1)
  } else {
    cacf_plot <- cacf_plot + geom_line(alpha = 0.5)
  }
  cacf_plot <- cacf_plot +
    theme_bw() +
    geom_hline(yintercept = 0) +
    scale_color_manual(values = vec_color) +
    facet_wrap(~ species, ncol = 2) +
    geom_vline(xintercept = year_marks, linetype = "dotted") +
    labs(
      #title = "Auto Correlation Function of TWD by Tree", 
      x = "Lag", y = toupper(my_y_val)) +
    theme(legend.position = "none")

}

# raw data
acf_data_list <- map(unique_tree_ids, get_acf)
acf_data <- bind_rows(acf_data_list)

acf_plot_by_species <- plot_corr_plots(my_df = acf_data)
ggsave(
  "images for report/acf_twd_by_species.png",
  plot = acf_plot_by_species, width = 10, height = 10
)

```

```{r corr_by_species_rolling7}
rolling.cols <- paste0(num.cols, "_rolling7")

for(iter_species in list_species){
  iter_df <- df.ts.series2 %>%
    filter(species == iter_species)

  print(iter_species)
  print(cor(iter_df[,rolling.cols], use = "complete.obs"))
}
```

```{r corr_by_species_rolling30}
rolling.cols <- paste0(num.cols, "_rolling30")

for(iter_species in list_species){
  iter_df <- df.ts.series2 %>%
    filter(species == iter_species)

  print(iter_species)
  print(cor(iter_df[,rolling.cols], use = "complete.obs"))
}
```

## Partial Autocorrelation
```{r}
# Plot PACF
get_pacf <- function(tid, y_val = "twd") {
  ts_data <- df.ts.series2 %>% filter(tree.id == tid)
  pacf_obj <- pacf(ts_data[[y_val]], lag.max = 2000, plot = FALSE)
  df <- tibble(
    tree_index = unique(ts_data$tree_index),
    species= unique(ts_data$species),
    tree.id = tid,
    lag = pacf_obj$lag[,1,1],
    pacf = pacf_obj$acf[,1,1]
  ) %>%
    filter(lag > 0)  # Exclude lag 0
  return(df)  
}
pacf_data_list <- map(unique_tree_ids, get_pacf)
pacf_data <- bind_rows(pacf_data_list)
pacf_plot_by_species <- plot_corr_plots(my_df = pacf_data, my_y_val = "pacf")

last_lag <- 100
pacf_plot_by_species_zoomed <- plot_corr_plots(pacf_data %>% filter(lag < last_lag), my_y_val = "pacf", add_smooth = FALSE) +
  coord_cartesian(xlim = c(0, last_lag))  # Zoom in on first 100 lags
ggsave(
  "images for report/pacf_twd_by_species_zoom.png",
  plot = pacf_plot_by_species_zoomed, width = 10, height = 10
)
```


```{r}
plot_twd_lag <- function(df, x_col, twd_col, lag = 1) {

  # Arrange data by time
  df <- df %>% arrange(.data[[x_col]])

  # Create lagged TWD values
  df <- df %>%
    mutate(Lagged_TWD = lag(.data[[twd_col]], lag))

  # Plot TWD vs. lagged TWD
  ggplot(df, aes(x = ts, color = tree.id, y = .data[[twd_col]])) +
    geom_line() +
    geom_line(aes(y= Lagged_TWD))+
    labs(title = paste("Tree Water Deficit vs. Lagged TWD (Lag =", lag, ")"),
         x = paste("TWD at Lag", lag),
         y = "TWD at Current Time") +
    theme_bw() +
    facet_wrap(~species)
}

plot_twd_lag(df.ts.series2, "at", "twd", lag = 5)

```


# Lagged correlation

```{r}
# Compute lagged correlation between variables and TWD
compute_lagged_corr <- function(df, var, twd_col = "twd", max_lag = 30) {
  map_dfr(1:max_lag, function(lag_val) {
    df_lagged <- df %>%
      mutate(var_lagged = lag(.data[[var]], lag_val)) %>%
      filter(!is.na(var_lagged), !is.na(.data[[twd_col]]))
    cor_val <- cor(df_lagged[[twd_col]], df_lagged$var_lagged, use = "complete.obs")
    tibble(variable = var, lag = lag_val, correlation = cor_val)
  })
}

#vec_covariates_short <- c("at_rolling30", "pr_rolling30")
vec_covariates_short <- c("at_rolling30")

# Compute lagged correlation by tree.id
lagged_corr_by_tree <- map_dfr(unique(df.ts.series2$tree.id), function(tid) {
  df_tree <- df.ts.series2 %>% filter(tree.id == tid)
  map_dfr(vec_covariates_short, ~ compute_lagged_corr(df_tree, .x, "twd_rolling30", max_lag = 100) %>%
            mutate(tree.id = tid))
})
# Add species information to lagged_corr_by_tree
lagged_corr_by_tree <- lagged_corr_by_tree %>%
  left_join(df.ts.series2 %>% select(tree.id, species) %>% distinct(), by = "tree.id")
# Plot lagged correlations by tree
ggplot(lagged_corr_by_tree, aes(x = lag, y = correlation, color = tree.id)) +
  geom_line(size = 0.8) +
  facet_wrap(~ species, scales = "free_y") +
  theme_bw() +
  labs(title = "Lagged Correlation between Covariates and TWD by Tree",
       x = "Lag (days)", y = "Correlation") +
  theme(legend.position = "bottom")

```



# Granger Causality

Granger Causality: Rather than testing whether X causes Y, the Granger causality tests whether X forecasts Y. 

A time series X is said to Granger-cause Y if it can be shown, usually through a series of t-tests and F-tests on lagged values of X (and with lagged values of Y also included), that those X values provide statistically significant information about future values of Y. 

A variable X that evolves over time Granger-causes another evolving variable Y if predictions of the value of Y based on its own past values and on the past values of X are better than predictions of Y based only on Y's own past values.  

# Test for stationarity

If the variables are non-stationary, then the test is done using first (or higher) differences. The number of lags to be included is usually chosen using an information criterion

```{r message = FALSE, warning = FALSE}
library(vars)
library(urca)

summary(ur.df(df.ts.series2$rolling30, type="drift", lags=2))

length(is.na(df.ts.series2$at_rolling30))/(40*30)
length(is.na(df.ts.series2$twd_rolling30))/(40*30)


# Test stationarity for at_rolling30 by tree.id
list_trees <- unique(df.ts.series2$tree.id)[1:3]
for (tid in list_trees) {
  cat("\nTree ID:", tid, "\n")
  sub_df <- df.ts.series2 %>% 
    filter(tree.id == tid) %>% 
    filter(!is.na(at_rolling30))

  print(summary(ur.df(sub_df$at_rolling30, type = "drift", lags = 2)))
  print(ggplot(sub_df, aes(x = ts, y = at_rolling30)) +
    geom_line() +
    labs(title = paste("Time Series of at_rolling30 for Tree ID", tid),
         x = "Time", y = "at_rolling30") +
    theme_minimal() +
    theme(legend.position = "none"))
}


```

Stationarity vs. seasonality

Stationarity requires constant mean and variance over time.

But in practice, deterministic seasonality violates this visually, because the mean changes with the time of year.

The ADF test does not explicitly model seasonal cycles, so it may flag the residual series as stationary even when the seasonality is strong.

Interpretation for your case:
The ADF test results indicate that the smoothed air temperature series does not have a stochastic trend (no random walk component), but the plot shows deterministic seasonal structure. That means the series is trend-stationary with strong seasonality, not strictly stationary in the sense of having a constant mean.

For your report, you could phrase it like this:
Although the Augmented Dickey–Fuller test strongly rejects the null of a unit root, indicating statistical stationarity, the rolling temperature series clearly exhibits deterministic seasonal cycles. This highlights that the test detects absence of stochastic trends, but does not remove or account for seasonality, which must be handled explicitly in modeling.

This tests the null hypothesis: The series has a unit root (i.e., non-stationary).

If the coefficient on z.lag.1 is significantly different from zero (with a very small p-value), then you can reject the null hypothesis and conclude the series is stationary.

# Granger


```{r run_granger_causality}

granger_test <- function(data, group = NA, x_col, y_col, lags = 10) {
  run_test <- function(df_sub) {
    df <- df_sub[, c(x_col, y_col)]
    names(df) <- c("x", "y")
    df <- df[complete.cases(df), ]
    
    if (nrow(df) <= lags + 1) {
      return("Not enough observations for VAR model.")
    }
    
    var_model <- VAR(df, p = lags, type = "const")
    return(causality(var_model, cause = "x"))
  }
  
  if (is.na(group)) {
    return(run_test(data))
  } else {
    group_list <- split(data, data[[group]])
    results <- lapply(group_list, run_test)
    return(results)
  }
}

granger_results <- list()
vec_covariates_roll <- paste0(vec_covariates, "_rolling30")
for(iter_covariate in vec_covariates_roll){
  granger_results[[iter_covariate]] <- granger_test(df.ts.series2, y_col= "twd_rolling30", x_col= iter_covariate, group = "tree.id")
}
```

```{r summarize_granger_results}
# Summarize Granger test p-values in a table (tree.id x covariate)
extract_pval <- function(granger_result) {
  print(head(granger_result))
  # Each element is a list for a tree, possibly a causality object or error string
  sapply(granger_result, function(res) {
      res$Granger$p.value
  })
}

tree_ids <- unique(df.ts.series2$tree.id)
covariate_names <- vec_covariates_roll

# Build p-value matrix
pval_mat <- sapply(covariate_names, function(cov) {
  vals <- extract_pval(granger_results[[cov]])
  # Ensure names match tree_ids
  vals[match(tree_ids, names(vals))]
})
rownames(pval_mat) <- tree_ids

# Convert to data frame for easier viewing
pval_df <- as.data.frame(pval_mat)
pval_df$tree.id <- rownames(pval_df)

# Reorder columns: tree.id first
pval_df <- pval_df[, c("tree.id", covariate_names)]

# Convert p-values to TRUE/FALSE for significance (e.g., p < 0.05)
sig_level <- 0.05
pval_df_tf <- pval_df
pval_df_tf[covariate_names] <- lapply(pval_df_tf[covariate_names], function(x) !is.na(x) & x < sig_level)
print(pval_df_tf)

# Print table
print(pval_df)
```

## Granger test by species

```{r}
granger_test(df.ts.series2, y_col= "twd", x_col= "at", group = "species")
```

## Linear regression

```{r}
plot_coeffs <- function(my_coef_df){
  ggplot(my_coef_df, aes(x = reorder(term, estimate), y = estimate)) +
    geom_bar(stat = "identity", fill = "steelblue") +
    geom_errorbar(aes(ymin = estimate - std.error, ymax = estimate + std.error), width = 0.2) +
    coord_flip() +  # flip axes for readability
    labs(
      x = "Predictor",
      y = "Estimate"
    ) +
    theme_minimal()
}
```

```{r regular_lm_all_species}
lm_fit <- lm(
  formula = as.formula(paste("twd_rolling30 ~ ", paste(x_vars, collapse = " + "))),
  data = df.ts.series2
)

x_vars <- c("at", "dp", "pr", "ws", "sr", "lr")
summary(lm_fit)
coef_df <- tidy(lm_fit)  # gets estimate, std.error, etc.
plot_lm_coeffs <- plot_coeffs(coef_df)
ggsave("images for report/plot_lm_coeffs.png", plot = plot_lm_coeffs, width = 4, height = 3)

x_vars_rolling7 <- c("at_rolling7", "dp_rolling7", "pr_rolling7", "ws_rolling7", "sr_rolling7", "lr_rolling7")
lm_fit_roll7 <- lm(
  formula = as.formula(paste("twd_rolling30 ~ ", paste(x_vars_rolling7, collapse = " + "))),
  data = df.ts.series2
)
summary(lm_fit_roll7)
coef_df <- tidy(lm_fit_roll7)  # gets estimate, std.error, etc.
plot_lm_coeffs <- plot_coeffs(coef_df)
ggsave("images for report/plot_lm_coeffs_rolling7.png", plot = plot_lm_coeffs, width = 4, height = 3)


x_vars_rolling30 <- c("at_rolling30", "dp_rolling30", "pr_rolling30", "ws_rolling30", "sr_rolling30", "lr_rolling30")
lm_fit_roll30 <- lm(
  formula = as.formula(paste("twd_rolling30 ~ ", paste(x_vars_rolling30, collapse = " + "))),
  data = df.ts.series2
)
summary(lm_fit_roll30)
coef_df <- tidy(lm_fit_roll30)  # gets estimate, std.error, etc.
plot_lm_coeffs <- plot_coeffs(coef_df)
ggsave("images for report/plot_lm_coeffs_rolling30.png", plot = plot_lm_coeffs, width = 4, height = 3)
```  

Series specific

```{r}
lm_fit <- lm(
  formula = twd ~ cum.time.days + at + dp + pr + ws + sr + lr,
  data = df.ts.series2,
  subset = species %in% c("Picea abies")
)

summary(lm_fit)
coef_df <- tidy(lm_fit)  # gets estimate, std.error, etc.
plot_coeffs(coef_df)
```  

```{r}
# Choose a random tree index from "Picea abies"
set.seed(36) # for reproducibility
picea_tree_indexes <- unique(df.ts.series2$tree_index[df.ts.series2$species == "Picea abies"])
random_picea_tree_index <- sample(picea_tree_indexes, 1)
cat("Random tree index from Picea abies:", random_picea_tree_index, "\n")

lm_fit_roll30_PiAbies <- lm(
  formula = as.formula(paste("twd_rolling30 ~ ", paste(x_vars_rolling30, collapse = " + "))),
  data = df.ts.series2,
  subset = tree_index == random_picea_tree_index
)

coef_df <- tidy(lm_fit_roll30_PiAbies) 
plot_coeffs(coef_df)

ggsave("images for report/annex/lin_reg_coeffs_pisceaabes.png", plot = plot_lm_coeffs, width = 4, height = 3)


Fagus_tree_indexes <- unique(df.ts.series2$tree_index[df.ts.series2$species == "Fagus sylvatica"])
random_fagus_tree_index <- sample(Fagus_tree_indexes, 1)
cat("Random tree index from Fagus sylvatica:", random_fagus_tree_index, "\n")

lm_fit_roll30_PiAbies_Fagus <- lm(
  formula = as.formula(paste("twd_rolling30 ~ ", paste(x_vars_rolling30, collapse = " + "))),
  data = df.ts.series2,
  subset = tree_index == random_fagus_tree_index
)

coef_df <- tidy(lm_fit_roll30_PiAbies_Fagus)
plot_coeffs(coef_df)

ggsave("images for report/annex/lin_reg_coeffs_fagussilv.png", plot = plot_lm_coeffs, width = 4, height = 3)

```

### Linear regression with autoregressive component

```{r}
# Linear regression with autoregressive component by tree.id

# Add lagged TWD columns (lags 1 to 10) for each tree
for (lag_num in 1:10) {
  df.ts.series2 <- df.ts.series2 %>%
    group_by(tree.id) %>%
    arrange(cum.time.days) %>%
    mutate(
      !!paste0("twd_lag", lag_num) := lag(twd, lag_num),
      !!paste0("twd_rolling30_lag", lag_num) := lag(twd_rolling30, lag_num)) %>%
    ungroup()
}

lag_cols <- paste0("twd_lag", 1:10)
# Fit linear model with lagged TWD for each tree
# Fit linear model with lagged TWD for each tree and extract R-squared
lm_ar_by_tree_model <- df.ts.series2 %>%
  group_by(tree.id) %>%
  nest() %>%
  mutate(
    model = map(data, ~ lm(as.formula(paste("twd ~ ", paste(x_vars, collapse = " + "), " +", paste(lag_cols, collapse = " + "))), data = .x)),
    coef = map(model, tidy),
    rsq = map_dbl(model, ~ summary(.x)$r.squared)
  )

lm_ar_by_tree <- lm_ar_by_tree_model %>%
  unnest(coef) %>%
  left_join(
    df.ts.series2 %>% dplyr::select(tree.id, species) %>% distinct(),
    by = "tree.id"
  ) %>%
  mutate(significant = ifelse(p.value < 0.05, "Yes", "No"))

# Plot AR(1) coefficient for each tree
ggplot(lm_ar_by_tree , aes(
  x = term, y = estimate)) +
  geom_jitter(aes(color = species, shape = significant), size = 2, width = 0.2, height =0) +
  geom_boxplot(aes(group = term), alpha = 0.2, outlier.shape = NA) +
  labs(x = "Predictor", y = "Coefficients", color = "Species", shape = "Is coefficient significant?") +
  theme_bw() +
  scale_color_manual(values = vec_color) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

ggsave("images for report/lin_reg_ar1_coeffs_by_tree.png", width = 10, height = 5)
```

```{r}
# Calculate mean R-squared per species
rsq_by_species <- lm_ar_by_tree %>%
  filter(term == "twd_lag1" ) %>%
  group_by(species) %>%
  dplyr::select(tree.id, species, rsq, estimate) %>%
  unique()

# Plot R-squared per species
ggplot(rsq_by_species, aes(x = species, y = rsq, color = estimate)) +
  geom_point(width = 0.2, alpha = 0.7, size = 2) +
  labs(x = "Species", y = "R-squared", title = "R-squared of AR Linear Model by Species") +
  theme_bw() +
  scale_color_gradient(low = "blue", high = "red") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

summary(rsq_by_species)
```

# Rollins SD

Confirm that timeseries are not stationary.

```{r fig.height =10}
# Extract residuals from the model column in lm_ar_by_tree_model
lm_ar_by_tree_model2 <- lm_ar_by_tree_model %>%
  mutate(
    fitted_values = map(model, fitted), 
    residuals = map(model, residuals))

# Example: view residuals for first tree
head(lm_ar_by_tree_model2$residuals[[1]])

plot_residuals <- function(df, x_col = "time_index") {
  # Plot residuals vs fitted for each tree
  print(ggplot(df, aes(time_index, residuals)) +
    geom_point(alpha = 0.6) +
    geom_hline(yintercept = 0, color = "red") +
    facet_wrap(~ tree.id, ncol = 4) +
    labs(x = "Cumulative Time (days)") +
    geom_vline(xintercept = year_marks, linetype = "dotted") +
    theme_bw()) +
    theme(axis.title.y = element_blank())
}

list_resid_plots <- list()
for(iter_model in 1:nrow(lm_ar_by_tree_model2)){
  df_resid <- tibble(
    time_index = seq( 1, length(lm_ar_by_tree_model2$residuals[[iter_model]])),
    tree.id = lm_ar_by_tree_model2$tree.id[iter_model],
    fitted_values = lm_ar_by_tree_model2$fitted_values[[iter_model]],
    residuals = lm_ar_by_tree_model2$residuals[[iter_model]]
  )
  list_resid_plots[[iter_model]] <- plot_residuals(df_resid)
}

indexes_to_view <- c(10, 20, 30, 40)
plots_residuals_to_show <- list_resid_plots[indexes_to_view]
ggsave("images for report/residuals_vs_time_index.png", 
       plot = marrangeGrob(plots_residuals_to_show, nrow=1, ncol=4), 
       width = 10, height = 4)

```